---
title: "CMIPB_batch_QC_correct_2020_2021"
output: html_notebook
author: Anna Konstorum (anna.konstorum@yale.edu)
---

# Install packages
```{r}
# If BatchQC not installed:
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("BatchQC")

# Note - I found that to get it to install, I had to first do
install.packages('seriation')
```

### Import libraries
```{r}
library(BatchQC)
library(sva)
```


### Load imputed data
Data source: [https://github.com/joreynajr/cmi-pb-multiomics/tree/main/results/main/cmi_pb_datasets/processed/imputed]  
Data was imputed using code in: [https://github.com/CMI-PB/2020_2021_data_merge/tree/main/scripts/CMIPB_2020_2021_imputation.Rmd]
Obtain files below from data source and save to local drive 
```{r}
# imputed_dir is path to local drive where data is saved
setwd(imputed_dir)
# import imputed datasets
rnaseq_baseline_mat_imputed_20 <- read.csv('rnaseq_baseline_mat_imputed_20_051022.csv',row.names=1)
cytof_baseline_mat_imputed_20 <- read.csv('cytof_baseline_mat_imputed_20_051022.csv',row.names=1)
olink_baseline_mat_imputed_20 <- read.csv('olink_baseline_mat_imputed_20_051022.csv',row.names=1)
abtiters_baseline_mat_imputed_20 <- read.csv('abtiters_baseline_mat_imputed_20_051022.csv',row.names=1)


rnaseq_baseline_mat_imputed_21 <- read.csv('rnaseq_baseline_mat_imputed_21_051022.csv',row.names=1)
cytof_baseline_mat_imputed_21 <- read.csv('cytof_baseline_mat_imputed_21_051022.csv',row.names=1)
olink_baseline_mat_imputed_21 <- read.csv('olink_baseline_mat_imputed_21_051022.csv',row.names=1)
abtiters_baseline_mat_imputed_21 <- read.csv('abtiters_baseline_mat_imputed_21_051022.csv',row.names=1)
```

`BatchQC` is a package that allows assessment of data for batch effects.  
For more information, see [https://academic.oup.com/bioinformatics/article/32/24/3836/2525651]
`ComBat` is a method to correct for batch effects, see[https://rdrr.io/bioc/sva/man/ComBat.html].  
There are many different options for Combat, depending on the distribution of your data.  
To help assess whether and how to use ComBat, we've created a helpful flow chart, [https://raw.githubusercontent.com/CMI-PB/2020_2021_data_merge/main/figures/BatchCorrection_Flowchart.png]

## Batch QC for cell frequency
### Prepare data
```{r}
# Data merge
cytof_all<-cbind(t(cytof_baseline_mat_imputed_20),t(cytof_baseline_mat_imputed_21))
batch<-c(rep('2020',nrow(cytof_baseline_mat_imputed_20)),rep('2021',nrow(cytof_baseline_mat_imputed_21)))
```

```{r}
cytof_20<-t(cytof_baseline_mat_imputed_20)
cytof_21<-t(cytof_baseline_mat_imputed_21)
mcytof_20<-melt(cytof_20)
mcytof_21<-melt(cytof_21)
mcytof_20$year<-"2020"
mcytof_21$year<-"2021"
mcytof<-rbind(mcytof_20,mcytof_21)
colnames(mcytof)<-c("cell_type","subject","value","year")
```

### Kolmogorov-Smirnoff test comparing each cell freq pop. dist. from 2020 v 2021
```{r}
k.s<-data.frame(matrix(nrow=nrow(cytof_20),ncol=3))
rownames(k.s)<-rownames(cytof_20)
colnames(k.s)<-c('mean_20','mean_21','ks_test')
for (i in 1:nrow(cytof_20)){
    cell_20<-cytof_20[i,]
    cell_21<-cytof_21[i,]
    k.s[i,"ks_test"]<-ks.test(cell_20,cell_21)$p.value
    k.s[i,'mean_20']<-mean(cell_20)
    k.s[i,'mean_21']<-mean(cell_21)
}
k.s$mean_20<-round(k.s$mean_20,3)
k.s$mean_21<-round(k.s$mean_21,3)
k.s$ks_test<-signif(k.s$ks_test,digits=3)
k.s<-k.s[order(nrow(k.s):1),]
```

### Visualize different distributions
```{r}
ggplot(mcytof, aes(x = value, y = cell_type)) +
  geom_density_ridges(aes(x=value, y=cell_type,
                      group = interaction(year,cell_type),
                      fill=year,alpha=0.5)) +
  theme_ridges() 
```
### Perform BatchQC
```{r}
batchQC(cytof_all, batch=batch,
        report_file="batchqc_report_Cytof_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

### Run ComBat (non-parametric) for cell frequency
Run non-parametric ComBat since data is not normally distributed and is not count data
```{r}
cytof_adjust<-ComBat(dat=cytof_all, batch=batch, par.prior=FALSE)
```

### Run BatchQC on adjusted data
```{r}
batchQC(cytof_adjust, batch=batch,
        report_file="batchqc_report_Cytof_adjust_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

## Batch QC for Olink
```{r}
# Olink merge
olink_all<-cbind(t(olink_baseline_mat_imputed_20),t(olink_baseline_mat_imputed_21))
batch<-c(rep('2020',nrow(olink_baseline_mat_imputed_20)),rep('2021',nrow(olink_baseline_mat_imputed_21)))
```

```{r}
batchQC(olink_all, batch=batch,
        report_file="batchqc_report_Olink_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

### Run ComBat on Olink data
```{r}
olink_adjust<-ComBat(dat=olink_all, batch=batch)
```
### Run BatchQC on adjusted data
```{r}
batchQC(olink_adjust, batch=batch,
        report_file="batchqc_report_Olink_adjust_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

## BatchQC for Ab titers
```{r}
# Olink merge
abtiters_all<-cbind(t(abtiters_baseline_mat_imputed_20),t(abtiters_baseline_mat_imputed_21))
batch<-c(rep('2020',nrow(abtiters_baseline_mat_imputed_20)),rep('2021',nrow(abtiters_baseline_mat_imputed_21)))
```

```{r}
batchQC(abtiters_all, batch=batch,
        report_file="batchqc_report_Abtiters_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```
Evaluation of batchQC output for abtiters shows no/few batch effects, so no batch corrections is needed.

## BatchQC for RNASeq
```{r}
# RNAseq merge
seqtiters_all<-cbind(t(rnaseq_baseline_mat_imputed_20),t(rnaseq_baseline_mat_imputed_21))
batch<-c(rep('2020',nrow(rnaseq_baseline_mat_imputed_20)),rep('2021',nrow(rnaseq_baseline_mat_imputed_21)))
```

```{r}
batchQC(seqtiters_all, batch=batch,
        report_file="batchqc_report_RNASeq_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```

### Run ComBat on RNASeq data
RNASeq data has been log-transformed, so can run ComBat (assuming near-normality of distribution)
```{r}
seqtiters_adjust<-ComBat(dat=seqtiters_all, batch=batch)
```

### Run BatchQC on adjusted data
```{r}
batchQC(seqtiters_adjust, batch=batch,
        report_file="batchqc_report_RNASeq_adjust_2020_2021.html", report_dir=".", 
        report_option_binary="111111111",
        view_report=FALSE, interactive=FALSE, batchqc_output=TRUE)
```
